---
title: "CRC Immune Phenotype and Prognostic Signature Pipeline"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: cosmo
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

options(stringsAsFactors = FALSE)

dir.create("data", showWarnings = FALSE)
dir.create("data/tcga", showWarnings = FALSE)
dir.create("data/geo", showWarnings = FALSE)
dir.create("results", showWarnings = FALSE)
dir.create("figures", showWarnings = FALSE)
```

# 1. Load packages

```{r packages}
needed_cran <- c(
  "dplyr", "ggplot2", "tibble", "stringr", "pheatmap",
  "reshape2", "data.table", "survival", "survminer",
  "glmnet", "readr"
)

for (pkg in needed_cran) {
  if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)
}

needed_bioc <- c(
  "TCGAbiolinks", "SummarizedExperiment", "GEOquery", "Biobase",
  "limma", "edgeR", "DESeq2", "clusterProfiler", "org.Hs.eg.db",
  "enrichplot", "GSVA"
)

if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")

for (pkg in needed_bioc) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    BiocManager::install(pkg, ask = FALSE, update = FALSE)
  }
}

library(dplyr)
library(ggplot2)
library(tibble)
library(stringr)
library(pheatmap)
library(reshape2)
library(data.table)
library(survival)
library(survminer)
library(glmnet)
library(readr)

library(TCGAbiolinks)
library(SummarizedExperiment)
library(GEOquery)
library(Biobase)
library(limma)
library(edgeR)
library(DESeq2)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(GSVA)
```

# 2. Notes before running

This report assumes you already generated or downloaded the core CRC expression and clinical data.

Required objects/files for the survival-signature part:

- `results/DE_with_symbols.csv`
- expression matrix with **gene symbols as rownames** and **samples as columns**
- survival table with columns:
  - `sample`
  - `time`
  - `status`

If you already saved the expression and survival objects before, this report will load them automatically.

# 3. Load saved objects if available

```{r load-saved}
has_expr <- file.exists("results/crc_tpm_log_gene.rds")
has_surv <- file.exists("results/surv_df2.rds")

if (has_expr) crc_tpm_log_gene <- readRDS("results/crc_tpm_log_gene.rds")
if (has_surv) surv_df2 <- readRDS("results/surv_df2.rds")

cat("Expression object found:", has_expr, "\n")
cat("Survival object found:", has_surv, "\n")
```

# 4. Optional: create expression and survival objects

Put your earlier code here if you need to rebuild `crc_tpm_log_gene` and `surv_df2`.

The final required format should be:

- `crc_tpm_log_gene`: matrix/data.frame, genes x samples
- `rownames(crc_tpm_log_gene)`: gene symbols
- `colnames(crc_tpm_log_gene)`: sample IDs
- `surv_df2`: data.frame with `sample`, `time`, `status`

```{r check-required}
if (!file.exists("results/DE_with_symbols.csv")) {
  stop("Missing file: results/DE_with_symbols.csv")
}

de_sym <- read.csv("results/DE_with_symbols.csv", stringsAsFactors = FALSE)

req_cols <- c("symbol", "adj.P.Val", "logFC")
missing_cols <- setdiff(req_cols, colnames(de_sym))
if (length(missing_cols) > 0) {
  stop("DE_with_symbols.csv is missing columns: ", paste(missing_cols, collapse = ", "))
}

if (!exists("crc_tpm_log_gene")) {
  stop("Object 'crc_tpm_log_gene' not found. Recreate it or load results/crc_tpm_log_gene.rds")
}

if (!exists("surv_df2")) {
  stop("Object 'surv_df2' not found. Recreate it or load results/surv_df2.rds")
}
```

# 5. Save key objects immediately

This prevents errors after an R session restart.

```{r save-core}
saveRDS(crc_tpm_log_gene, "results/crc_tpm_log_gene.rds")
saveRDS(surv_df2, "results/surv_df2.rds")
```

# 6. Create strict DE gene list

```{r strict-de}
sig_strict_genes <- unique(
  na.omit(de_sym$symbol[de_sym$adj.P.Val < 0.01 & abs(de_sym$logFC) > 1.5])
)

cat("Strict genes (symbols):", length(sig_strict_genes), "\n")
```

# 7. Match genes and samples

```{r match-data}
genes_for_cox <- intersect(sig_strict_genes, rownames(crc_tpm_log_gene))
cat("Strict genes present in expr:", length(genes_for_cox), "\n")

if (length(genes_for_cox) == 0) {
  stop("No strict DE genes are present in the expression matrix.")
}

req_surv <- c("sample", "time", "status")
missing_surv <- setdiff(req_surv, colnames(surv_df2))
if (length(missing_surv) > 0) {
  stop("surv_df2 is missing columns: ", paste(missing_surv, collapse = ", "))
}

common_samples <- intersect(surv_df2$sample, colnames(crc_tpm_log_gene))
cat("Matched samples:", length(common_samples), "\n")

if (length(common_samples) == 0) {
  stop("No matching samples between surv_df2 and expression matrix.")
}

surv_df2_use <- surv_df2 %>%
  filter(sample %in% common_samples) %>%
  distinct(sample, .keep_all = TRUE) %>%
  filter(!is.na(time), !is.na(status))

surv_df2_use$time <- as.numeric(surv_df2_use$time)
surv_df2_use$status <- as.numeric(surv_df2_use$status)

surv_df2_use <- surv_df2_use %>%
  filter(!is.na(time), !is.na(status), time > 0)

cat("Samples after filtering:", nrow(surv_df2_use), "\n")
table(surv_df2_use$status)
summary(surv_df2_use$time)
```

# 8. Build expression matrix for Cox analysis

```{r build-x}
X_all2 <- t(crc_tpm_log_gene[genes_for_cox, surv_df2_use$sample, drop = FALSE])
X_all2 <- as.matrix(X_all2)

keep_no_na <- apply(X_all2, 2, function(z) all(!is.na(z)))
X_all2 <- X_all2[, keep_no_na, drop = FALSE]

sdv <- apply(X_all2, 2, sd)
X_all2 <- X_all2[, sdv > 0, drop = FALSE]

cat("Genes after NA/zero-variance removal:", ncol(X_all2), "\n")

if (ncol(X_all2) == 0) {
  stop("No genes left after filtering.")
}
```

# 9. Univariate Cox ranking

```{r univariate-cox}
y <- Surv(surv_df2_use$time, surv_df2_use$status)

cox_p <- apply(X_all2, 2, function(x) {
  fit <- tryCatch(coxph(y ~ x), error = function(e) NULL)
  if (is.null(fit)) return(NA_real_)
  out <- tryCatch(summary(fit)$coef[1, "Pr(>|z|)"], error = function(e) NA_real_)
  as.numeric(out)
})

cox_res <- data.frame(
  gene = names(cox_p),
  p = as.numeric(cox_p),
  stringsAsFactors = FALSE
) %>%
  filter(!is.na(p), is.finite(p)) %>%
  arrange(p)

write.csv(cox_res, "results/univariate_cox_ranked_p.csv", row.names = FALSE)
saveRDS(cox_res, "results/cox_res.rds")

head(cox_res, 10)
```

# 10. LASSO Cox model

```{r lasso-fit}
topN <- min(300, nrow(cox_res))
lasso_genes <- head(cox_res$gene, topN)
X_lasso <- X_all2[, lasso_genes, drop = FALSE]

cat("Genes entering LASSO:", ncol(X_lasso), "\n")

saveRDS(X_lasso, "results/X_lasso.rds")

set.seed(1)
cvfit <- cv.glmnet(
  x = X_lasso,
  y = y,
  family = "cox",
  alpha = 1,
  nfolds = 10
)

saveRDS(cvfit, "results/cvfit_custom.rds")
plot(cvfit)
```

# 11. Choose lambda for a 13-gene signature

```{r choose-lambda}
beta <- as.matrix(coef(cvfit$glmnet.fit))
n_genes <- colSums(beta != 0)

target <- 13
exact_idx <- which(n_genes == target)

if (length(exact_idx) > 0) {
  idx <- exact_idx[1]
} else {
  idx <- which.min(abs(n_genes - target))
}

lambda_custom <- cvfit$glmnet.fit$lambda[idx]
coef_custom <- coef(cvfit$glmnet.fit, s = lambda_custom)

nz <- which(as.numeric(coef_custom) != 0)

signature_custom <- data.frame(
  gene = rownames(coef_custom)[nz],
  coef = as.numeric(coef_custom)[nz],
  stringsAsFactors = FALSE
) %>%
  arrange(desc(abs(coef)))

cat("Final signature size (custom):", nrow(signature_custom), "\n")
signature_custom
```

# 12. Save signature

```{r save-signature}
write.csv(signature_custom, "results/lasso_signature_lambda_custom.csv", row.names = FALSE)
saveRDS(signature_custom, "results/signature_custom.rds")
```

# 13. Compute risk score

```{r risk-score}
sig_genes <- intersect(signature_custom$gene, rownames(crc_tpm_log_gene))
cat("Signature genes found in expression:", length(sig_genes), "\n")

X_sig <- t(crc_tpm_log_gene[sig_genes, surv_df2_use$sample, drop = FALSE])
X_sig <- as.matrix(X_sig)

coef_df <- signature_custom[match(colnames(X_sig), signature_custom$gene), ]
beta_sig <- coef_df$coef

risk_score <- as.numeric(X_sig %*% beta_sig)

surv_sig <- surv_df2_use
surv_sig$risk_score <- risk_score
surv_sig$risk_group <- ifelse(
  surv_sig$risk_score >= median(surv_sig$risk_score, na.rm = TRUE),
  "High", "Low"
)

write.csv(surv_sig, "results/risk_scores.csv", row.names = FALSE)

head(surv_sig)
```

# 14. Kaplan-Meier analysis

```{r km-plot, fig.width=8, fig.height=6}
fit_km <- survfit(Surv(time, status) ~ risk_group, data = surv_sig)

p_km <- ggsurvplot(
  fit_km,
  data = surv_sig,
  pval = TRUE,
  risk.table = TRUE,
  conf.int = FALSE,
  title = "Kaplan-Meier Survival by Risk Group"
)

p_km
```

# 15. Optional: Cox model using risk score

```{r cox-risk}
cox_risk <- coxph(Surv(time, status) ~ risk_score, data = surv_sig)
summary(cox_risk)
```

# 16. Session info

```{r session-info}
sessionInfo()
```